---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# HCAmtxzip

<!-- badges: start -->
<!-- badges: end -->

This package is not mature.

This package implements tools to access and manipulate Human Cell
Atlas '.mtx.zip' archives describing single cell expression
studies. The focus is on the pre-computed matrices available by
navigating projects on the Human Cell Atlas Data Portal [project
summary page][hca_projects].

[hca_projects]: https://data.humancellatlas.org/explore/projects

## Installation

To install this package and it's dependencies, use BiocManager

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cran.r-project.org")

BiocManager::install("mtmorgan/HCAmtxzip")
```

Load the package and discover project titles for which pre-computed
mtx.zip archives are available

```{r discover}
suppressPackageStartupMessages({ library(HCAmtxzip) })
dd = discover()
dd
```

Find the smallest archive, and pull the path to the mtx.zip archive

```{r}
path <- dd %>%
    filter(row_number() == which.min(size)) %>%
    pull(path)
```

Download, unzip, and import the archive into _R_; the download is
cached (using [BiocFileCache][]) so only necessary once. The imported
data is represented as a [SingleCellExperiment][], a data structure at
the heart of _Bioconductor_ packages [orchestrating single cell
analysis][osca]. The `dim` output shows that there are 58347 features
assayed across 2544 cells.

[BiocFileCache]: https://bioconductor.org/packages/BiocFileCache
[SingleCellExperiment]: https://bioconductor.org/packages/SingleCellExperiment
[osca]: https://osca.bioconductor.org

```{r import}
sce <- import.mtxzip(path, verbose = TRUE)
sce
```

The figure below provides a schematic view of the SingleCellExperiment
-- there is feature x cell matrix of counts, with row and column
annotations.

![sce][sce_image]

[sce_image]: https://raw.githubusercontent.com/Bioconductor/SummarizedExperiment/master/vignettes/SE.svg?sanitize=true

The object contains a matrix of feature x sample counts, and it's easy
to discover, e.g., that about 92\% the cells in the matrix are zeros.

```{r}
mean(assay(sce) == 0)
```

Information about each feature can be extracted with `rowData(sce)`,
and about each cell with `colData(sce)`. Much of the `colData()` is
constant across all cells, e.g., they all come from the same
experiment. A helper function `colData_common()` summarizes these
common features, where we learn for instance that the experiment
involves the pancreas islet of Langerhans.

```{r}
colData_common(sce) %>% print(n=Inf)
```

The helper function `colData_distinct()` extracts the columns that
differ between cells

```{r}
colData_distinct(sce)
```

An additional helper function shortens column names to the shortest
'word' (defined by the occurence of `_` or `.`) suffix, as illustrated
with

```{r}
x <- setNames(nm = c("common_prefix.a_title", "common_prefix.a_name"))
names_abbreviate(x)
names_abbreviate(x, map = TRUE)
```

This can make the column names a little easier to use in an
interactive session.

```{r}
colData_distinct(sce, abbreviate_colnames = TRUE)
```

Some exploration suggests that the column
`donor_organism.provenance.document_id` defines each biological
sample, so we see that the experiment consists of 8 individuals with
between 178 and 505 cells per individual.

```{r}
colData_distinct(sce, TRUE) %>%
    count(donor_organism.provenance.document_id)
```

There are 6 male and 2 female samples with diverse ethnicity.

```{r}
colData_distinct(sce, TRUE) %>%
    count(donor_organism.provenance.document_id, sex, ethnicity.ontology_label)
```

The provenance `document_id` serves as a link to addition information
about the samples. The additional infomration can be extracted by
querying the HCA DSS. The information is returned as JSON, which is
relatively difficult to deal with in _R_; the even-less-mature
function `.files()` tries to parse this into a sensible format.

```{r donor_files}
uuids <- colData_distinct(sce, TRUE) %>%
    count(donor_organism.provenance.document_id) %>%
    pull(donor_organism.provenance.document_id)
donor <- HCAmtxzip:::.files(uuids)
donor
```

We thus learn the sex, age and cause of death of each individual

```{r}
donor %>%
    select(provenance.document_id, sex, organism_age, death.cause_of_death)
```

Similarly detailed information about other aspects of the experiment,
such as contributors, funding, and publications, can be learned by
querying `.files()` with additional document identifiers from
elsewhere in `colData(sce)`. For instance, here's the (truncated, in
the HCA) description of the project.

```{r project_files}
uuid <- colData_common(sce) %>% 
    filter(key %in% "project.provenance.document_id") %>% 
    pull(value)
project <- HCAmtxzip:::.files(uuid)
project
pull(project, "project_core.project_description") %>%
    strwrap(width = 80) %>%
    cat(sep="\n")
```

Here's information about the software in use to produce the above.

```{r sessionInfo}
sessionInfo()
```
