---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
```

# HCAmtxzip

<!-- badges: start -->
<!-- badges: end -->

This package is not mature.

This package implements tools to access and manipulate Human Cell
Atlas '.mtx.zip' archives describing single cell expression
studies. The focus is on the pre-computed matrices available by
navigating projects on the Human Cell Atlas Data Portal [project
summary page][hca_projects].

[hca_projects]: https://data.humancellatlas.org/explore/projects

## Installation

To install this package and its dependencies, use BiocManager

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cran.r-project.org")

BiocManager::install("mtmorgan/HCAmtxzip")
```

## Discovery and import

Load the package and find available project titles for which
pre-computed mtx.zip archives are available

```{r available}
suppressPackageStartupMessages({ library(HCAmtxzip) })
dd <- available()
dd
```

Find the smallest archive

```{r}
dd %>%
    filter(size == min(size))
```

Download, unzip, and import the archive into _R_; the download is
cached (using [BiocFileCache][]) so the download is only necessary
once. The imported data is represented as a [SingleCellExperiment][],
a data structure at the heart of _Bioconductor_ packages
[orchestrating single cell analysis][osca]. The `dim` output shows
that there are 58347 features assayed across 2544 cells.

[BiocFileCache]: https://bioconductor.org/packages/BiocFileCache
[SingleCellExperiment]: https://bioconductor.org/packages/SingleCellExperiment
[osca]: https://osca.bioconductor.org

```{r import}
sce <- 
    filter(dd, size == min(size)) %>%
    import.mtxzip(path, verbose = TRUE)
sce
```

The figure below provides a schematic view of the SingleCellExperiment
-- there is feature x cell matrix of counts, with row and column
annotations.

![sce][sce_image]

[sce_image]: https://raw.githubusercontent.com/Bioconductor/SummarizedExperiment/master/vignettes/SE.svg?sanitize=true

The object contains a matrix of feature x sample counts, and it is easy
to find, e.g., that about 92\% the cells in the matrix are zeros.

```{r}
mean(assay(sce) == 0)
```

Information about each feature can be extracted with `rowData(sce)`,
and about each cell with `colData(sce)`; it is often convenient to work
with this data using the 'tidy' framework, with `rowDataTibble()` and
`colDataTibble()` providing relevant access

```{r}
colDataTibble(sce)
```

Much of the `colData()` is constant across all cells, e.g., all cells
come from the same experiment, so share the same
`project.project_core.project_title`. Use `colDataConstants()` to
access these common features, where we learn for instance that the
experiment involves the pancreas islet of Langerhans.

```{r}
colDataConstants(sce) %>%
    print(n = Inf)
```

The function `colDataBrief()` extracts the columns that differ between
cells, and shortens column names to the shortest 'word' (delimited by
`[[:punct:]]+`) suffix, as illustrated with

```{r}
x <- setNames(nm = c("common_prefix.a_title", "common_prefix.a_name"))
HCAmtxzip:::.names_abbreviate(x)
```

This can make the column names a little easier to use in an
interactive session.

```{r}
colDataBrief(sce)
```

## Working with colData

Some exploration suggests that the column
`donor_organism.provenance.document_id` defines each biological
sample, so we see that the experiment consists of 8 individuals with
between 178 and 505 cells per individual.

```{r}
colDataBrief(sce) %>%
    count(donor_organism.provenance.document_id)
```

There are 6 male and 2 female samples with diverse ethnicity.

```{r}
count <-
    colDataBrief(sce) %>%
    count(donor_organism.provenance.document_id, sex, ethnicity.ontology_label)
count
```

The `document_iddonor_organism.provenance.document_id` serves as a
link to addition information about the samples. The additional
information can be extracted by querying the HCA DSS. The information
is returned as JSON, which is relatively difficult to deal with in
_R_; the even-less-mature function `.files()` tries to parse this into
a sensible format.

```{r donor_files}
donor <-
    colDataTibble(sce) %>%
    distinct(donor_organism.provenance.document_id) %>%
    HCAmtxzip:::.files(donor_organism.provenance.document_id)
```

We thus learn the age and cause of death of each individual

```{r}
donor %>%
    select(provenance.document_id, sex, organism_age, death.cause_of_death)
```

The information on donor could be joined with the count summary, or
with the entire data set using `left_join()`

```{r}
left_join(count, donor)
all <- left_join(colDataTibble(sce), donor)
```

Similarly detailed information about other aspects of the experiment,
such as contributors, funding, and publications, can be learned by
querying `.files()` with additional document identifiers from
elsewhere in `colData(sce)`. For instance, here is the (truncated, in
the HCA) description of the project.

```{r project_files}
project <- colDataTibble(sce) %>%
    distinct(project.provenance.document_id) %>%
    HCAmtxzip:::.files(project.provenance.document_id)
project
pull(project, "project_core.project_description") %>%
    strwrap(width = 80) %>%
    cat(sep="\n")
```

## Exploratory analysis

To illustrate the first analytic steps enabled by this package, load
the [scater][] (single-cell normalization) and [ggplot2][]
(visualization) packages.

[scater]: https://bioconductor.org/packages/scater
[ggplot2]: https://cran.r-project.org/package=ggplot2

```{r, message = FALSE}
library(scater)
library(ggplot2)
```

Normalize our data; the updated `sce` contains an addition assay `logcounts`.


```{r}
sce <- logNormCounts(sce) 
```

For an overall gestalt, focus on the 1000 most variable
(statistically informative?) genes


```{r rowvar}
rv = apply(logcounts(sce), 1, var)
keep <- tail(order(rv), 1000)
```

and perform a principle components analysis

```{r prcomp}
pc = prcomp(t(logcounts(sce)[keep,]), scale = TRUE)
```

Visualize the results using the [ggplot2][] package, where each point
represents the log-normalized expression of a cell, reduced to
two-dimensional space.

```{r prcomp_plot}
ggplot(as_tibble(pc$x), aes(PC1, PC2)) + geom_point()
```

It then becomes interesting to identify the factors that separate the
cells into groups.

## Acknowledgements

A portion of this work is supported by the Chan Zuckerberg Initiative
DAF, an advised fund of Silicon Valley Community Foundation.

Research reported in this presentation was supported by the NHGRI and
NCI of the National Institutes of Health under award numbers
U41HG004059, U24CA180996, and U24CA232979. The content is solely the
responsibility of the authors and does not necessarily represent the
official views of the National Institutes of Health.


The following is a summary of software used to produce this document.

```{r sessionInfo}
sessionInfo()
```
